---
title: "Bayesian generalized linear modelling"
date: "31st March 2022"
output:
  word_document: default
  fig_width: 5 
  fig_height: 5 
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(runjags)

```  

# Forest avifauna exhibit enduring responses to historical high-severity wildfires  

The number of bird species in relation to fire predictors in montane dry sclerophyll forests.    

*Study area*    
Greater Blue Mountains World Heritage Area, Australia.  
The 48 sites were distributed along an ~ 80 km latitudinal gradient from the upper Blue Mountains to Gospers Mountain in the north.       

*Site level*    
All bird surveys were conducted in late spring/early summer 2018. Nocturnal species were excluded from the bird data.    

LAT_s; standardized (mean = 0, s.d. = 1) northing (latitude) at the center of each site.     
MFC_s; standardized area-weighted mean number of fires that burnt each site between 1982 and 2018.    
FY2; an indicator variable for year of last fire (0 if 2002, 1 if 2013).       
SEV; an indicator variable for high-severity fire, where 0 = less than half the site had been burned by at least one high-severity fire since 1982, and 1 = more than 50% of the site was burnt by such a fire.    
INT; variable for the FY2*SEV interaction.    


### Read in the data

```{r}

# Read in the binary bird response data - species not detected (0) or detected (1) in each site.   
# The order of rows and columns correspond to site numbers and species numbers respectively (ascending order).  
path <- here("data", "bird_data3.csv")
dat.counts <- read.csv(path, header = FALSE, stringsAsFactors = FALSE)

# Site attributes, predictor variables.
path <- here("data", "sites_lookup3.csv")
SitesLookup <- read.csv(path, stringsAsFactors = FALSE)

# Species numbers, codes and names.
path <- here("data", "species_lookup3.csv")
SpeciesLookup <- read.csv(path, stringsAsFactors = FALSE)

```


### Count the number of species recorded in each site

```{r}

No.species <- cbind(1:48, rowSums(dat.counts))
colnames(No.species)[1:2] <- c("site", "Sp")

No.species <- as.data.frame(No.species)

No.species <- No.species %>% left_join(SitesLookup, by = "site")

DAT <- No.species %>%
  dplyr::select(nspecies = Sp, FY2, SEV, INT, LAT_s, MFC_s)

```


### Check for over- or under-dispersion by fitting a simple GLM with quasipoisson family

```{r}

m_quas <- glm(nspecies ~ LAT_s + MFC_s + FY2 + SEV + INT, family = quasipoisson(link = "log"), data = DAT)
summary(m_quas)

# The data are under-dispersed (estimated dispersion ~ 0.49). 

```


### JAGS model code    
Equation 2     

Michael Bedward adapted the model code below from an example in:      
Hilbe, J.M., De Souza, R.S., Ishida, E.E., 2017. Bayesian models for astrophysical data: using R, JAGS, Python, and Stan. Cambridge University Press.        

See: https://www.bayesianmodelsforastrophysicaldata.com/code-6-18-and-6-19 for original code.    

```{r}

GPmodel_code <- "model {
  # Priors for regression coefficients
  beta0 ~ dnorm(0, 0.1)
  betaFY2 ~ dnorm(0, 0.1)
  betaSEV ~ dnorm(0, 0.1)
  betaINT ~ dnorm(0, 0.1)
  betaLAT ~ dnorm(0, 0.1)
  betaMFC ~ dnorm(0, 0.1)

  # Prior for delta parameter of GP distribution
  delta ~ dunif(-1, 1)

  # Constant for 'zeros trick'
  C <- 10000
  
  for (i in 1:Nsites){
    Zeros[i] ~ dpois(Zeros.mean[i])
    Zeros.mean[i] <- -L[i] + C
    
    l1[i] <- log(mu[i])
    l2[i] <- (nspecies[i] - 1) * log(mu[i] + delta * nspecies[i])
    l3[i] <- -mu[i] - delta * nspecies[i]
    l4[i] <- -loggam(nspecies[i] + 1)
    L[i] <- l1[i] + l2[i] + l3[i] + l4[i]
    
    mu[i] <- (1 - delta)*exp(loglinpredictor[i])
    loglinpredictor[i] <- beta0 + betaLAT * LAT_s[i] + betaMFC * MFC_s[i] + betaFY2 * FY2[i] + 
                          betaSEV * SEV[i] + betaINT * INT[i]
  }

  # Discrepancy measures: mean, variance, Pearson residuals
  for (i in 1:Nsites){
    ExpY[i] <- mu[i] / (1 - delta)
    VarY[i] <- mu[i] / ((1 - delta)^3)
    Pres[i] <- (nspecies[i] - ExpY[i]) / sqrt(VarY[i])
  } 
  
}"

```


### Model data

```{r}

dat_model <- c(as.list(DAT), list(Nsites = nrow(DAT), Zeros = rep(0, nrow(DAT))))

```


### Run the model

```{r}

GPmodel <- run.jags(GPmodel_code, monitor = c("beta0", "betaLAT", "betaMFC", "betaFY2", "betaSEV", "betaINT", "delta"),   
                    data = dat_model, 
                    n.chains = 4,
                    sample = 2000,
                    thin = 10)

```


### Model summaries and plots

```{r}

summary(GPmodel)

plot(GPmodel)

```

